Concurrent Programming in Harmony

### 2020-12-27 book/paper.tex
Latest

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 53 diameter = 5
==== Safety violation ====
T0: __init__() [0-4,31-46] 46 { count: 0, done[0]: False, done[1]: False }
T1: incrementer(0) [5-8]    9 { count: 0, done[0]: False, done[1]: False }
T2: incrementer(1) [5-15]  15 { count: 1, done[0]: False, done[1]: True  }
T1: incrementer(0) [9-15]  15 { count: 1, done[0]: True,  done[1]: True  }
T3: main()         [17-27] 27 { count: 1, done[0]: True,  done[1]: True  }
>>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
Open file:///Users/rvr/github/harmony/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/Up1.pdf}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are three sections.
The top shows the steps to the problematic state.
The bottom left shows the source code and bytecode
of the program.
The bottom right shows the state of each thread after the selected step.
Each row in the stack trace of a thread shows the current program counter,
the method that is being evaluated, and the line of code that is
being evaluated.}
\label{fig:inchtml1}
\end{figure}

@@

The first step is initialization.
It sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
If we look at the next three steps, we see that:
\begin{itemize}
\item Thread \texttt{incrementer/0}
executed instructions 5 through 8, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer/1}
executed instructions 5 through 15, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer/0}
continues execution, executing instructions 9 through 15
storing value 1 into \textit{count}
and storing \texttt{True} into \textit{done}[0].
\end{itemize}
Finally, thread \texttt{main()} runs and finds the problem.
This makes precise the concurrency issue that we encountered.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
\hline
program counter & an integer value pointing into the code \\
\hline
frame pointer & an integer value pointing into the stack \\
\hline
atomic & if non-zero, the thread is in atomic mode \\
\hline
readonly & if non-zero, the thread is in read-only mode \\
\hline
stack & a list of Harmony values \\
\hline
register & a dictionary mapping atoms (names of variables) to values \\
\hline
stopped & a boolean indicating if the context is stopped \\
\hline
failure & if not None, string that describes how the thread failed \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The name in a name tag is the name of the method that the thread
is executing;
\item The tag in a name tag the argument to the method by default.  Optionally
\texttt{spawn} allows the tag to be specified explicitly;
\item The frame pointer points to the current \emph{stack frame},
which consists of the caller's frame pointer and variables, the argument to
the method, an ``invocation type atom'' (\texttt{normal}, \texttt{interrupt}, or
\texttt{thread}), and the return address (in case of \texttt{normal}).
\item A thread terminates when it reaches the \texttt{Return} instruction
of the top-level method (when the stack frame is of type \texttt{thread})
or when it hits an exception.  Exceptions include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, infinite loops of microsteps, as well as when an assertion fails;
\item The execution of a thread in \emph{atomic mode} does not get interleaved
with that of other threads.
\item The execution of a thread in \emph{read-only mode} is not allowed
to update shared variables of spawn threads.
\item The register of a thread always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a Harmony program.
\end{itemize}

@@

There is initially a single context with nametag
\texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
in atomic mode until it finishes executing the last instruction in
the code.  All states until then are \texttt{initializing} states.
Other threads, created through \texttt{spawn} statements, do not
start executing until then.

>>>>

\begin{figure}
\begin{code}
\begin{verbatim}
#states = 53 diameter = 5
==== Safety violation ====
T0: __init__() [0-4,31-46] 46 { count: 0, done[0]: False, done[1]: False }
T1: incrementer(0) [ 5- 8]  9 { count: 0, done[0]: False, done[1]: False }
T2: incrementer(1) [ 5-15] 15 { count: 1, done[0]: False, done[1]: True  }
T1: incrementer(0) [ 9-15] 15 { count: 1, done[0]: True,  done[1]: True  }
T3: main()         [17-27] 27 { count: 1, done[0]: True,  done[1]: True  }
>>> Harmony Assertion (file=code/Up.hny, line=10) failed: 1
Open file:///Users/rvr/github/harmony/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/Up1.png}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are three sections.
The top shows the steps to the problematic state.
The bottom left shows the source code and bytecode
of the program.
The bottom right shows the state of each thread after the selected step.
Each row in the stack trace of a thread shows the current program counter,
the method that is being evaluated, and the line of code that is
being evaluated.}
\label{fig:inchtml1}
\end{figure}

@@

The first step is initialization.
It sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
If we look at the next three steps, we see that:
\begin{itemize}
\item Thread \texttt{incrementer(0)}
executed instructions 5 through 8, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer(1)}
executed instructions 5 through 15, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer(0)}
continues execution, executing instructions 9 through 15
storing value 1 into \textit{count}
and storing \texttt{True} into \textit{done}[0].
\end{itemize}
Finally, thread \texttt{main()} runs and finds the problem.
This makes precise the concurrency issue that we encountered.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name & the name of the main method that the thread is executing \\
\hline
argument & the argument given to the main method \\
\hline
program counter & an integer value pointing into the code \\
\hline
frame pointer & an integer value pointing into the stack \\
\hline
atomic & if non-zero, the thread is in atomic mode \\
\hline
readonly & if non-zero, the thread is in read-only mode \\
\hline
stack & a list of Harmony values \\
\hline
method variables & a dictionary mapping atoms (names of method variables) to values \\
\hline
thread-local variables & a dictionary mapping atoms (names of thread-local variables) to values \\
\hline
stopped & a boolean indicating if the context is stopped \\
\hline
failure & if not None, string that describes how the thread failed \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The frame pointer points to the current \emph{stack frame},
which consists of the caller's frame pointer and variables, the argument to
the method, an ``invocation type atom'' (\texttt{normal}, \texttt{interrupt}, or
\texttt{thread}), and the return address (in case of \texttt{normal}).
\item A thread terminates when it reaches the \texttt{Return} instruction
of the top-level method (when the stack frame is of type \texttt{thread})
or when it hits an exception.  Exceptions include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, infinite loops of microsteps, as well as when an assertion fails;
\item The execution of a thread in \emph{atomic mode} does not get interleaved
with that of other threads.
\item The execution of a thread in \emph{read-only mode} is not allowed
to update shared variables of spawn threads.
\item The register of a thread always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a Harmony program.
\end{itemize}

@@

There is initially a single context with name
\texttt{\_\_init\_\_/()} and program counter 0.  It starts executing
in atomic mode until it finishes executing the last instruction in
the code.  All states until then are \texttt{initializing} states.
Other threads, created through \texttt{spawn} statements, do not
start executing until then.
